Motion planning for assessing accessibility of protein tunnels

### 2017-09-10 paper.tex
image of dcp trajectories + tunnel from Adam
added colormap form Adam
atom-based metric

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{zhang2008fast,he2016efficient}. %\cite{he2016efficient} in RAL!!
Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
the retractions only to the narrow passages.
Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed toward the medial axis.
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

@@

In this paper, we propose a modification of the RRT planner for computing trajectories through protein tunnels.
The input is a set of tunnels computed by Voronoi diagrams and the task is to find trajectories along the given tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence of protein dynamics that is computed using MD simulations.

@@

The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of the dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it support retraction of the ligand to the narrow passages.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.

A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
The tunnels can be found by the tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$ and the actual scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
The traversability of the tunnel should therefore be evaluated using various starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

\subsection{RRT for computing ligands trajectory inside a tunnel}

The proposed motion planning is based on the principle of RRT~\cite{lavalleRRT} planner.
The main loop of planner (Alg.~\ref{alg::main}) is similar to the original RRT planner.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
%This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
This allows the tree to slightly detour from the tunnel, which improves performance around narrow tunnels.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause  exploration of paths around the tunnel.
We propose to set this parameter to the average tunnel width.

@@

The second proposed modification of RRT is the expansion procedure (Alg.~\ref{alg::expand}), which generates new collision-free nodes around $\qnear\in\T$.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qrand$ and tested for collisions.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.

The third modification is the utilization of another metric during the expansion procedure.
While the nearest neighbors are search using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
procedure utilizes a metric that is sensitive to the shape of the ligands.
This is necessary as the ligands adopt various shapes according to the actual conformations.
To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.

Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration
$q$ being tested.
This metric is denoted $\da(\cdot,\cdot)$ on line~\ref{alg::expand:a} in Alg.~\ref{alg::expand} and it is referred to as atom-based metric in the rest of the paper.
As the expansion routine attempts to find such sample that minimizes this distance to $\qrand$, it supports retraction of the ligand towards $\qrand$.
The fast evaluation of this metric with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, can be achieved using KD-trees.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find nearest point towards $\qrand'$, which
is also the nearest neighbor of the ligand at $q$ to $\qrand$.

@@

>>>>

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
the retractions only to the narrow passages.
Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on e.g. random vectors, obstacles vectors or medial axis in order to improve growth in the narrow passages.
In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed toward the medial axis.
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

@@

We propose a modification of the RRT planner for computing trajectories through protein tunnels.
The input is a set of tunnels computed using Voronoi diagrams and the task is to find trajectories along the given tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence of protein dynamics that is computed using MD simulations.

@@

The ligand flexibility is modeled using a predefined set of typical conformations, therefore the number of the dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
Library of conformations are used also in different types of calculations, e.g.~\cite{kellogg}. 
While classic 6D Euclidean metric is used in the nearest-neighbor search of RRT, we propose to use another metric during the expansion step.
This metric measures distances between two configurations as the nearest distance of two atoms. 
This enables the metric to consider the shape of the ligand and it supports retraction of the ligand to the narrow passages.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
A protein tunnel is described by a sequence of collision-free spheres 
$T=( (c_1, r_1),\ldots,(c_n,r_n) )$, where $n$ denotes the number of spheres,
$c_i \in \R^3$ is their 3D position and $r_i > 0$ denotes the maximum collision-free radius of a sphere centered at $c_i$. 
The tunnels can be found by the tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (the geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$ and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
To find a new initial configuration, a random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation  parts of $q$ are generated randomly, the scale is set to $\smin$ and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

\subsection{RRT for computing ligands trajectory along a tunnel}

The proposed motion planning is based on the principle of the RRT~\cite{lavalleRRT} planner.
In each iteration, a random sample $\qrand$ is generated and its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between $\qrand$ and the tree is performed using the weighted 6D Euclidean metric computed using
both 3D rotation and 3D translation.
The main loop of the planner (Alg.~\ref{alg::main}) is similar to original RRT, but we propose three modifications.

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This allows the tree to slightly detour from the tunnel and overcome its difficult (narrow) parts by finding an alternative trajectory nearby.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
of $\rv$ cause  exploration of paths around the tunnel.
We propose to set this parameter to the average width of the tunnel.

@@

The second proposed modification is the expansion procedure (Alg.~\ref{alg::expand}), which expands the tree from $\qnear$ towards $\qrand$.
The expansion towards $\qrand$ is crucial for the growth of the tree towards unexplored areas of the configuration space.
Traditionally, a line is constructed from $\qnear$ to $\qrand$ and the tree is expanded by a configuration located on this line.
This would however mostly lead to a collision in the case of ligands.

Instead, we propose to expand the tree by considering all conformations and also by expanding the tree using the maximal possible scale that
does not lead to a collision.
For each ligand conformation $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with a maximal scale.
First, the maximal scale $\smax \in \S$ is selected and $m$ random samples are generated around $\qnear$ and tested for collisions.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
The random samples are generated similarly as in the case of $\qrand$ samples, only their translation
part $(x,y,z)$ is generated around $\qnear$.

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
is useful when multiple conformations are utilized.
By employing this metric in the expansion procedure, the tree is expanded by such a new configurations, whose atoms maximally approach
the atoms of the configuration $\qrand$.
We observed that this leads to a better construction of the configuration tree inside the proteins and it increases probability
of reaching the active site.

%While the nearest neighbors are searched using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
%procedure utilizes a metric that is sensitive to the shape of the ligands.
%This is necessary as the ligands adopt various shapes according to the actual conformations.
%To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.

%Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
The atom-based metric can be evaluated fast with complexity $O(\log n)$, where $n$ is the number of atoms of the ligand, using KD-trees.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $T=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-T)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
is also the nearest point of the ligand at $q$ to $\qrand$.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{fig/dcp-image1Label}\\
\includegraphics[width=0.2\textwidth]{fig/colormap}
\caption{\label{fig::dcp}
Visualization of trajectories for the ligand DCP   in the two examined tunnels.  
    The color of the trajectories corresponds to the scale, that was needed to pass the given part of the tunnel.
    For the purpose of this image, also scales $\smin=0.3$ and $\smin=0.4$ where tested.
}
\end{figure}
